---
title: "Analysis of uplift RCT"
output: html_document
date: "2026-05-02"
author: Tilman Bayer
---

# Libraries and settings
```{r}
library(googlesheets4)
library(data.table)
library(clubSandwich)
```


```{r}
# Seaborn "colorblind" palette, copied from SEABORN_PALETTES in:
# https://github.com/mwaskom/seaborn/blob/master/seaborn/palettes.py
# Seaborn documents "colorblind" as a named palette:
# https://seaborn.pydata.org/generated/seaborn.color_palette.html
colorblind_palette <- c(
  "#0173B2",
  "#DE8F05",
  "#029E73",
  "#D55E00",
  "#CC78BC",
  "#CA9161",
  "#FBAFE4",
  "#949494",
  "#ECE133",
  "#56B4E9"
)
```

# Ingest data from Google Sheets
```{r echo=FALSE}
# NB: This may require authorization with your Google account in interactive mode 
# first (reply to the question in the console) before using it while knitting
d <- read_sheet('https://docs.google.com/spreadsheets/d/1bEDYkg4epbnJr5JUJ-P10KVICxhoYQhGx2CXVOOGCf8/edit')
# TODO: replace with a reference to the published version of the data once released
```

```{r}
# inspect data:
head(d)
```


```{r}
dt <- data.table(d)
names(dt)
```

```{r}
# Retrieve column name for session duration (since it has a complicated name, 
# inherited from the wording of this question in Google Forms):
duration_col <- grep("duration", names(dt), ignore.case = TRUE, value = TRUE)
if (length(duration_col) != 1) {
  stop("Expected exactly one duration column, found: ", paste(duration_col, collapse = ", "))
}

durs <- dt[[duration_col]]
```


## Histogram of session durations
```{r}
hist_breaks <- seq(
  floor(min(durs, na.rm = TRUE) / 15) * 15,
  ceiling(max(durs, na.rm = TRUE) / 15) * 15,
  by = 15
)
hist_xlim <- range(c(hist_breaks, 180))
hist_tick_at <- seq(
  floor(hist_xlim[1] / 30) * 30,
  ceiling(hist_xlim[2] / 30) * 30,
  by = 30
)
hist_tick_at <- hist_tick_at[hist_tick_at >= hist_xlim[1] & hist_tick_at <= hist_xlim[2]]
hist_axis_at <- sort(unique(c(hist_tick_at, hist_xlim)))
hist_axis_labels <- ifelse(hist_axis_at %in% hist_tick_at, hist_axis_at, "")
```


```{r duration-by-condition}
condition_col <- "Condition"

duration_by_condition <- split(durs, dt[[condition_col]])
duration_by_condition <- lapply(duration_by_condition, function(x) x[!is.na(x)])
duration_by_condition <- duration_by_condition[lengths(duration_by_condition) > 0]

condition_colors <- rep(colorblind_palette, length.out = length(duration_by_condition))

hist_data <- lapply(
  duration_by_condition,
  hist,
  breaks = hist_breaks,
  plot = FALSE
)
hist_colors <- grDevices::adjustcolor(
  condition_colors,
  alpha.f = 0.45
)
hist_ylim <- c(0, max(unlist(lapply(hist_data, function(x) x[["counts"]]))))

par(family = "serif")
for (i in seq_along(hist_data)) { # loop through conditions
  plot(
    hist_data[[i]], # plot the ith histogram object
    col = hist_colors[i],
    border = "white",
    xlab = "minutes",
    ylab = "",
    main = "Duration of reproduction sessions",
    # font size magnification factors:
    cex.main = 1.875,
    cex.axis = 1.04,
    xlim = hist_xlim,
    ylim = hist_ylim,
    xaxt = "n", # suppress the default x-axis
    add = i > 1 # create plot if i=1, add plot if i>1
  )
}

# add custom x axis, and y axis label:
axis(1, at = hist_axis_at, labels = hist_axis_labels, cex.axis = 1.2)
mtext("number of sessions", side = 2, line = 3, cex = 1.04)

legend(
  "top",
  legend = names(duration_by_condition),
  fill = hist_colors,
  border = "white",
  cex = 1.5
)
```

## Uplift analysis

This regression estimates the association between condition (AI-assisted vs. Manual) and log duration. The underlying experiment was designed assuming that both researchers (evaluators) and papers (replication targets) may have unobserved characteristics that affect task duration. This motivates the use of a blocked randomization assignment (blocking on both researchers and papers), and the use of fixed effects for both in the model.

In the paper's notation, the model is:

$$
\log(\mathrm{duration}_{i}) =
\alpha
+ \beta \, \mathrm{AI}_{i}
+ \gamma_p
+ \delta_r
+ \varepsilon_i,
$$

Here, $i$ indexes the individual reproduction session, $p$ denotes the paper whose result is being reproduced in $i$, and $r$ the researcher conducting the session. The terms $\gamma_p$ are paper fixed effects, $\delta_r$ are researcher fixed effects, and $\beta$ is the estimated uplift from AI assistance relative to the manual condition.

```{r prepare data for FE models}
condition_col <- "Condition"
reference_condition <- "AI-assisted" # estimate change relative to this condition
# Uplift is generally conceived as a speed change in this context,
# and speed is reciprocal to duration. So for simplicity we estimate 
# the reciprocal factor for duration (from AI-assisted to Manual instead of vice versa).

paper_col <- "Paper Title"
researcher_col <- "Human Researcher [your name]"

fixed_effects_regression_data <- data.frame(
  duration_minutes = durs,
  condition = relevel(factor(dt[[condition_col]]), ref = reference_condition),
  paper = factor(dt[[paper_col]]),
  researcher = factor(dt[[researcher_col]])
)
```

```{r FE model}
m_fe <- lm(
  log(duration_minutes) ~ condition + paper + researcher,
  data = fixed_effects_regression_data
)

condition_effect_term <- grep(
  "^condition",
  names(coef(m_fe)),
  value = TRUE
)
if (length(condition_effect_term) != 1) {
  stop("Expected exactly one condition effect, found: ", paste(condition_effect_term, collapse = ", "))
}

condition_label <- sub("^condition", "", condition_effect_term)

# Use the actual model frame to avoid potential row-alignment problems:
mf <- model.frame(m_fe)
```

```{r CR2 clustered by researcher}
V_researcher_cr2 <- vcovCR(
  m_fe,
  cluster = mf$researcher,
  type = "CR2"
)

researcher_cr2_condition_test <- coef_test(
  m_fe,
  vcov = V_researcher_cr2,
  test = "Satterthwaite",
  coefs = condition_effect_term
)
researcher_cr2_condition_log_effect <- researcher_cr2_condition_test$beta
researcher_cr2_condition_std_error <- researcher_cr2_condition_test$SE
researcher_cr2_condition_df <- researcher_cr2_condition_test$df_Satt
researcher_cr2_condition_p_value <- researcher_cr2_condition_test$p_Satt
researcher_cr2_condition_multiplier <- exp(researcher_cr2_condition_log_effect)
researcher_cr2_condition_percent_change <-
  100 * (researcher_cr2_condition_multiplier - 1)
researcher_cr2_condition_variance <- researcher_cr2_condition_std_error^2

researcher_cr2_condition_test

```


Using CR2 standard errors clustered by researcher, the coefficient for `r condition_label` is `r round(researcher_cr2_condition_log_effect, 4)`, with a CR2 standard error of `r round(researcher_cr2_condition_std_error, 4)`, Satterthwaite degrees of freedom of `r round(researcher_cr2_condition_df, 1)`, and p-value `r format.pval(researcher_cr2_condition_p_value, digits = 3)`. The point estimate implies that `r condition_label` sessions last `r round(researcher_cr2_condition_multiplier, 2)` times as long as `r reference_condition` sessions, or about `r round(researcher_cr2_condition_percent_change, 0)` percent as long.


```{r CR2 clustered by paper}
V_paper_cr2 <- vcovCR(
  m_fe,
  cluster = mf$paper,
  type = "CR2"
)

paper_cr2_condition_test <- coef_test(
  m_fe,
  vcov = V_paper_cr2,
  test = "Satterthwaite",
  coefs = condition_effect_term
)
paper_cr2_condition_log_effect <- paper_cr2_condition_test$beta
paper_cr2_condition_std_error <- paper_cr2_condition_test$SE
paper_cr2_condition_df <- paper_cr2_condition_test$df_Satt
paper_cr2_condition_p_value <- paper_cr2_condition_test$p_Satt
paper_cr2_condition_multiplier <- exp(paper_cr2_condition_log_effect)
paper_cr2_condition_percent_change <- 100 * (paper_cr2_condition_multiplier - 1)
paper_cr2_condition_variance <- paper_cr2_condition_std_error^2

paper_cr2_condition_test
```

Using CR2 standard errors clustered by paper, the coefficient for `r condition_label` is `r round(paper_cr2_condition_log_effect, 4)`, with a CR2 standard error of `r round(paper_cr2_condition_std_error, 4)`, Satterthwaite degrees of freedom of `r round(paper_cr2_condition_df, 1)`, and p-value `r format.pval(paper_cr2_condition_p_value, digits = 3)`. The point estimate is unchanged, but this version allows residual dependence within papers rather than within researchers.
